Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2014-01-30 paper.Rnw
Fixed FEP experiment evaluation and methods description

\subsubsection{Materials and Methods: \expadnixval{}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%     RESULTS: Experiment 1: Whole Hippocampus Cross-Validation       %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Results: \expadnixval{}}

We find that for \mb{} segmentations, similarity score increases as atlas and
template library size is increased, although with diminishing returns and an
eventual trend towards a plateau (Figure \ref{fig:ADNI1-xval-k-mean}).  A
maximum similarity score of \Sexpr{mb.kappa.best} is found when using 9 atlases,
19 templates, ANTS registration, and majority vote label fusion).  The \ants{}
registration method consistently out-performs \animal{} registration over all
variable settings we tested. Additionally, by itself, using a weighted voting
strategy did not significantly improve segmentation accuracy, contrary to the
findings of \citet{Aljabar2009} using basic \multiatlas{} segmentation.  Given
these findings, in the remainder of this section only results using the \ants{}
registration algorithm and majority vote fusion will be shown.

@@

We find similar behaviour with respect to optimal parameter settings and
increased consistency of \mb{} segmentations in the replication of this
experiment (\expadnixvalsnt{}, Supplementary Materials) where atlases use a different
hippocampal definition is used. This strongly suggests that these results are
independant of the segmentation protocol used and are, instead, features of the
\mb{} algorithm.

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 2: Application of \mb{} to the segmentation      %%%%%
%%%%%                       of Psychosis patients                         %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\expfepxval{}: \expfepxvaldesc{}}
\label{exp:fep-xval}
\textbf{[TODO: tie with exp 1]}
To validate that the \mb{} works effectively in the context of other neurological
disorders, in this experiment we use the Winterburn atlases to derive 
whole hippocampal segmentations of a dataset of patients having had a single
episode of psychosis. The resulting segmentations are assessed for quality by
comparison with expert manual segmentations.

\subsubsection{Materials and Methods: \expfepxval{}}
\paragraph{First Episode Psychosis Dataset}
All patients were recruited and treated through the Prevention and Early
Intervention Program for Psychoses (PEPP-Montreal), a specialized early
intervention service at the Douglas Mental Health University Institute in
Montreal, Canada. People aged 14 to 35 years from the local catchment area
suffering from either affective or non-affective psychosis who had not taken
antipsychotic medication for more than one month with an IQ above 70 were
consecutively admitted as either in- or out-patients. Of those treated at PEPP,
only patients aged 18 to 30 years with no previous history of neurological
disease or head trauma causing loss of consciousness were eligible for the
neuroimaging study; only those suffering from schizophrenia spectrum disorders
were considered for this analysis.  For complete program details see
\citet{Malla2003}.

@@

\paragraph{Experiment details} 
\mb{} is configured with an atlas library composed of the Winterburn T1 atlases
(see \expfepxval{}) ignoring subfield delineations.  All images from the
first episode dataset are segmented by \mb{}.  The optimal size of template
library, registration method, and label fusion method found in \expadnixval{} are
used.

\paragraph{Evaluation method}
The Pruessner and Winterburn hippocampal segmentation protocols differ slightly in
the neuroanatomical features that are delineated \citep{Winterburn2013}. This
difference poses a problem for evaluation by measuring overlap.  That is, since
different protocols will necessarily produce segmentations that do not perfectly
overlap, the degree of overlap cannot be solely used to compare segmentation
methods using different protocols. In place of an overlap metric, we can assess
the degree of (Pearson) correlation in average bilateral hippocampal volume of the
subjects produced by each method.

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%    RESULTS: Experiment 2, Application of \mb{} to the segmentation  %%%%%
%%%%%                       of psychosis patients                         %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Results: \expfepxval{}}
\textbf{[TODO: move up results to here.]}

@@

\subsubsection{Materials and Methods: \expadni{}}
\paragraph{\adnidataset{} dataset} 
The {\em \adnidataset{}} standardized dataset contains $1919$ images in total.
\snt{}, \maper, \freesurfer{} hippocampal volumes for a subset of images were
provided by ADNI, along with quality control data for each \freesurfer{}
segmentation (guidelines described in \citep{Hartig2010}).

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                           RESULTS                                   %%%%%
%%%%%         EXPERIMENT 3: Application of \mb{} to the segmentation      %%%%%
%%%%%                       of ADNI subjects                              %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Results: \expadni{}}

@@

\subsubsection{Materials and Methods: \expsubfield{}}
\paragraph{Winterburn Atlases} 
The Winterburn atlases \citep{Winterburn2013} are digital hippocampal
segmentations of five in-vivo $0.3mm$-isotropic T1-weighted MR images. The
segmentations include subfield segmentations for the cornu ammonis (CA) 1; CA2
and CA3; CA4 and dentate gyrus; subiculum; and strata radiatum (SR), strata
lacunosum (SL), and strata moleculare (SM). Subjects in the Winterburn atlases
range in age from 29-57 years (mean age of 37), and include two males and three
females.

\paragraph{Experiment details} 
Leave-one-out cross-validation (LOOCV) is a validation approach in which an
algorithm is given all but one item in a dataset as training data (in our case,
atlas images and labels) and then the algorithm is applied to the left-out item.
This is done, in turn, for each item in the dataset and the output across all
items is evaluated together.

@@

\paragraph{Evaluation method}  
Evaluating the accuracy of hippocampal subfields for T1 images at
$0.9mm^3$-isotropic voxels is inherently ill-defined since there are no manual
protocols for segmentation at this resolution to use as a reference. Instead, we
can evaluate the reliability, or {\it precision}, with which \mb{} produces
hippocampal subfields that correspond to the segmentation protocol used by the
given atlas library images. By directly resampling the Winterburn atlas
segmentations to $0.9mm^3$ voxels (using standard nearest-neighbour image
resampling techniques) we obtain a scaled down version of the labels which
preserve the original segmentation protocol within the limits of error from
rounding and interpolation. Therefore, using the resampled Winterburn
segmentations as definitive for the $0.9mm^3$ resolution we evaluate reliability
of \mb{} segmentations using using DSC overlap scores and evaluate consistency
across the range of hippocampal sizes using Bland-Altman plots of subfield
volumes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%    RESULTS: Experiment 4: Hippocampal Subfield Cross-Validation     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Results: \expsubfield{}}

@@

\subsubsection{Materials and Methods: \expadnixvalsnt{}}
\paragraph{Dataset} 
69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset{}} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Each image has a corresponding
semi-automated segmentation of the left and right whole hippocampus made
available with ADNI images (\snt; see Supplementary Materials).

@@

% ######################
\subsubsection{Results: \expadnixvalsnt{}}

>>>>

\subsubsection{\expadnixval{}: Materials and Methods}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%     RESULTS: Experiment 1: Whole Hippocampus Cross-Validation       %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{\expadnixval{}: Results}

We find that for \mb{} segmentations, similarity score increases as atlas and
template library size is increased, although with diminishing returns and an
eventual trend towards a plateau (Figure \ref{fig:ADNI1-xval-k-mean}).  A
maximum similarity score of \Sexpr{mb.kappa.best} is found when using 9 atlases,
19 templates, ANTS registration, and majority vote label fusion).  The \ants{}
registration method consistently out-performs \animal{} registration over all
variable settings we tested. Additionally, by itself, using a weighted voting
strategy did not significantly improve segmentation accuracy, contrary to the
findings of \citet{Aljabar2009} using basic \multiatlas{} segmentation.  Given
these findings, in the remainder of our experiments only results using the \ants{}
registration algorithm and majority vote fusion will be shown.

@@

We find similar behaviour with respect to optimal parameter settings and
increased consistency of \mb{} segmentations in the replication of this
experiment (\expadnixvalsnt{}, Supplementary Materials) where a different
hippocampal definition is used (\snt labels available with the ADNI datasets).
This strongly suggests that these results are independant of the segmentation
protocol used and are, instead, features of the \mb{} algorithm.

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 2: Application of \mb{} to the segmentation      %%%%%
%%%%%                       of Psychosis patients                         %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\expfepxval{}: \expfepxvaldesc{}}
\label{exp:fep-xval}
\textbf{[TODO: tie with exp 1]}
To validate that the \mb{} works effectively in the context of other neurological
disorders, in this experiment we use the Winterburn atlases
\citep{Winterburn2013} to derive whole hippocampal segmentations of a dataset of
patients having had a single episode of psychosis. The resulting segmentations
are assessed for quality by comparison with expert manual segmentations.

\subsubsection{\expfepxval{}: Materials and Methods}
\paragraph{First Episode Psychosis Dataset}
All patients were recruited and treated through the Prevention and Early
Intervention Program for Psychoses (PEPP-Montreal), a specialized early
intervention service at the Douglas Mental Health University Institute in
Montreal, Canada. People aged 14 to 35 years from the local catchment area
suffering from either affective or non-affective psychosis who had not taken
antipsychotic medication for more than one month with an IQ above 70 were
consecutively admitted as either in- or out-patients. Of those treated at PEPP,
only patients aged 18 to 30 years with no previous history of neurological
disease or head trauma causing loss of consciousness were eligible for the
neuroimaging study; only those suffering from schizophrenia spectrum disorders
were considered for this analysis.  For complete program details see
\citet{Malla2003}.

@@

\paragraph{Winterburn Atlases} 
The Winterburn atlases \citep{Winterburn2013} are digital hippocampal
segmentations of five in-vivo $0.3mm$-isotropic T1-weighted MR images. The
segmentations include subfield segmentations for the cornu ammonis (CA) 1; CA2
and CA3; CA4 and dentate gyrus; subiculum; and strata radiatum (SR), strata
lacunosum (SL), and strata moleculare (SM). Subjects in the Winterburn atlases
range in age from 29-57 years (mean age of 37), and include two males and three
females.

\paragraph{Experiment details} 
The same overall design as \expadnixval{} is followed in this experiment: a
Monte Carlo cross-validation (MCCV) is conducted using the pool of 81 first
episode psychosis subject brain images and corresponding \prsnr-protocol manual
segmentations. Five rounds of validation are conducted for each subject, and
each atlas and template library size combination (1-9 atlases, 1-19 templates).
In each round, images and their manual labels are randomly selected from the
pool, and the remaining images are segmented using \mb{} with a random subset of
the unlabelled images also serve as template images. Majority vote fusion, and
the \ants{} registration algorithm are used, as these have shown to behave
favourably in previous experiments.

In addition to the MCCV, we segment the entire first episode psychosis dataset
using \mb{} and two other popular segmentation packages, \fsl{} and
\freesurfer{}. Firstly, \mb{} is used with the five Winterburn atlas images and labels as atlases, and a
randomly selected subset of 19 psychosis dataset images as templates. Secondly,
\mb{} is run using the same template images, but instead as atlases we use five
additional five first episode psychosis subjects and corresponding manual
segmentations (not included above). Finally, \fsl{} and \freesurfer{} are run
with their default settings to segment the entire dataset. The \fsl
\verb+run_first_all+ script was used according to the FIRST user guide
\footnote{http://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide}, and
\freesurfer was run using \verb+recon-all+. \textbf{

\paragraph{Evaluation method}
Manual and automated segmentations are directly compared using Dice's similarity
coefficent (DSC). In the MCCV, the per-subject DSC value is computed as the
average value over the five rounds of validation for a given atlas and template
library size. The reported average DSC value per given atlas and template
library size is the average DSC value over all subjects segmented.

The Pruessner segmentation protocol differs slightly from the Winterburn
protocol those used by \freesurfer and \fsl in the neuroanatomical features that
are delineated (see \citet{Winterburn2013}, and Table
\label{tab:subfield-segmentation-protocols} in the Discussion below).  This
variation in protocol poses a problem if an overlap measure is used for
evaluation: since different protocols will necessarily produce segmentations
that do not perfectly overlap, the degree of overlap cannot be solely used to
compare segmentation methods using different protocols. In place of an overlap
metric, we assess the degree of (Pearson) correlation in average bilateral
hippocampal volume produced by each method. Additionally, we evaluate the
volume-related fixed and proportional biases in all segmentation methods using
Bland-Altman plots \citep{Bland1986}.

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%    RESULTS: Experiment 2, Application of \mb{} to the segmentation  %%%%%
%%%%%                       of psychosis patients                         %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{\expfepxval{}: Results}
\textbf{[TODO: move up results to here.]}

@@

\subsubsection{\expadni{}: Materials and Methods}
\paragraph{\adnidataset{} dataset} 
The {\em \adnidataset{}} standardized dataset contains $1919$ images in total.
\snt{}, \maper, \freesurfer{} hippocampal volumes for a subset of images were
provided by ADNI, along with quality control data for each \freesurfer{}
segmentation (guidelines described in \citep{Hartig2010}).

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                           RESULTS                                   %%%%%
%%%%%         EXPERIMENT 3: Application of \mb{} to the segmentation      %%%%%
%%%%%                       of ADNI subjects                              %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{\expadni{}: Results}

@@

\subsubsection{\expsubfield{}: Materials and Methods}
%^ TODO: winterburn atlases described in expfepxval
\paragraph{Experiment details} 
Leave-one-out cross-validation (LOOCV) is a validation approach in which an
algorithm is given all but one item in a dataset as training data (in our case,
atlas images and labels) and then the algorithm is applied to the left-out item.
This is done, in turn, for each item in the dataset and the output across all
items is evaluated together.

@@

\paragraph{Evaluation method}  
Evaluating the accuracy of hippocampal subfields for T1 images at
$0.9mm^3$-isotropic voxels is inherently ill-defined since there are no manual
protocols for segmentation at this resolution to use as a reference. Instead, we
can evaluate the reliability, or {\it precision}, with which \mb{} produces
hippocampal subfields that correspond to the segmentation protocol used by the
given atlas library images. By directly resampling the Winterburn atlas
segmentations to $0.9mm^3$ voxels (using standard nearest-neighbour image
resampling techniques) we obtain a scaled down version of the labels which
preserve the original segmentation protocol within the limits of error from
rounding and interpolation. Therefore, using the resampled Winterburn
segmentations as definitive for the $0.9mm^3$ resolution we evaluate reliability
of \mb{} segmentations using DSC overlap scores and evaluate consistency
across the range of hippocampal sizes using Bland-Altman plots of subfield
volumes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%    RESULTS: Experiment 4: Hippocampal Subfield Cross-Validation     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{\expsubfield{}: Results}

@@

\subsubsection{\expadnixvalsnt{}: Materials and Methods}
\paragraph{Dataset} 
69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset{}} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Each image has a corresponding
semi-automated segmentation of the left and right whole hippocampus made
available with ADNI images (\snt; see Supplementary Materials).

@@

% ######################
\subsubsection{\expadnixvalsnt{}: Results}
